Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-14 manuscript.Rnw
edits to text, attempt to better integrate climate effects and AF/AS

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GMVs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig. 2c and S3).

@@

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFTa dn plant height (and the quadratic term) as predictors (and all interactions). We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs. 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs. 0.28). This is surprising because LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

Given these results, we were also interested to know whether additional variation in biomass partitioning could be explained by either climate, within PFT variation in LMA or by other traits. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons among species and across climatic gradients, due to the small sample size within each species or site. Thus, to answer these additional questions we found it helpful to decompose LMF as the product,

@@

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and climates.

@@

\emph{I think the point is that we found MF/AS to be proportional to LMA across PFTs, however, the other component of MF/AS is AF/AS and that in fact, across species, explained more variation in MF/AS. So, though LMA is a good substitute for PFT, lots of other variation still exists, which should be taken into account. We just have shown that AF/AS does not depend on PFT, and almost not on climate - so it has to be some unidentified factor instead. ??}. In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning. The effect of LMA was found to hold also within PFTs (Fig. 4a), explaining 28\% of the variation in $M_F/A_S$ across species. The fact that PFTs differ in LMA is evident in the separation of the three groups along the x-axis of Fig. 4a. Yet these clouds were broad and overlapping, with a more general relationship extending across the entire LMA axis. The remaining (?? what is number?? \emph{the remainder should be 100 - 28}) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig. 4b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig. 2e), but does vary close to two orders of magnitude across species. The separation among PFTs along the Y-axis arises due to differences in LMA.

>>>>

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GVMs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = \textrm{LMF} / \textrm{LMA}$ (Fig. 2c and S3).

@@

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions). We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs. 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs. 0.28). This is surprising because LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}). We were thus interested to know whether additional variation in biomass partitioning could be explained by climate, as a basis for model formulation. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site. Thus, we decomposed LMF as the product,

@@

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables 1-2). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering which traits affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

\begin{equation}\label{eq:M2}
\frac{M_F}{A_S} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{A_S}\right).
\end{equation}

The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig. 4a), explaining 30\% of the variation in $M_F/A_S$ across species. The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis. A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig. 4b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig. 2e), but does vary close to two orders of magnitude across species. In Fig. 4b, The separation among PFTs in $M_F/A_S$ arises due to differences in LMA. We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
